Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-06-24 revision/review_response.Rmd

## Reviewer 2

### Overview

This manuscript presents a new application of GAMs approaches to fit irregular time series data (gaps, changing frequency of measurements) to estimate trends. The paper presents a case study using Chl a data from San Francisco Bay.  The manuscript is presented as a general approach with applications to other similar environmental time series data.  A comparison of methods shows that the GAMs plus meta-analysis method provides some different trend conclusions from ordinary least squares and plain GAMs. The paper is generally well written and well-suited for the audience for STOTEN.  I have some clarifications that would help improve/clarify the utility and readability of the manuscript.  These are presented below.

* __Response__: Thank you for your comment on our manuscript.  We have addressed your concerns on the utility and readability of our manuscript below.

@@

### Utility (1):

We applied the wqtrends code in R to a time series data set of X2 (salinity intrusion measure) over approximately in San Francisco Bay.  Some observations:

-the code did not work with the full time series of daily X2 data.  But we were able to make it work using a random sampling of 5 points per month.  Could the reason for this error be explored?  As time series data are concerned, we provided an input of ~100 years of daily data.  A sample result for the trends are shown in the figure below.  I think it would be helpful for readers/users to understand what the practical limitations of the code/algorithm are.

* __Response__: Very rarely do reviewers actually apply the proposed methods on novel datasets as a proof of concept.  For that, we are very appreciative of your efforts in exploring the R package. These applications are important for testing both the method and software.  As such, you exposed an important limitation of our method that was not clearly articulated in the draft.  We provide two responses to this limitation.

     First, the X2 dataset is collected daily, so the GAM smoothing methods proposed in our manuscript for "irregularly spaced or missing" monitoring data is unnecessary.  The modelling approach is designed to provide both a continuous (e.g., daily) estimate of the long-term trend and a measure of uncertainty in the absence of more regularly collected values.  So, this would be an unnecessary applications of GAMs.

     Second, and more important, the practical limitations of applying a GAM with flexible smoothing to nearly 100 years of daily data is beyond the computational power most desktop computers.  This is asking too much for the smoothing splines used in the mgcv package. Given this limitation, we do feel it is important to clearly state the intended use case for these models to avoid confusion in application.  We have added clarification both in the manuscript and the package vignette so users are clear on the type of data that is appropriate for the methods.

     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the the intended use case for the methods herein."

     In the package vignette, first paragraph (https://tbep-tech.github.io/wqtrends/articles/introduction.html): "These models are appropriate for data typically collected for surface water quality monitoring programs at roughly monthly or biweekly collection intervals, covering at least a decade of observations (e.g., Cloern and Schraga 2016). Daily or continuous monitoring data are not appropriate for these methods, due to computational limitations and a goal of the analysis to estimate long-term, continuous trends from irregular or discontinuous sampling."

@@

### Utility (2):

From figure 7 it seems that there are very different conclusions for Station 36 depending on the method used.  From Figure 8, however, it seems that Station 36 is a bit of an anomaly in this regard, and that other 8 stations in the case study have findings on trends that are similar among the three methods—and that the results are not dramatically different.  Perhaps this should be the basis for added discussion in the final section—when is the added complexity of the GAMS and meta analysis method is appropriate in a real-world setting?  Alternatively, it would be helpful if the authors were to examine another data set from the region that they work with, to review the relative findings of trend significance and what this depends on.

* __Response__: You are correct to note that many of the comparisons had similar results and we agree that it is useful to clarify that our approach is warranted relative to more conventional methods.  If a simpler approach is used, there is no way to know for sure that a false positive or negative result is obtained without applying the proposed methods.  So, we argue that using our approach is still the wiser option in all cases.  Otherwise, a user cannot be certain the correct conclusions are obtained. We have added the following to the discussion:

    Line xyz: "Moreover, use of methods that fully account for uncertainty in sampling design is recommended over more conventional analyses given the certainty of conclusions in the former.  Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used. Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences."

@@

The term meta-analysis seems to be an important part of the paper and is referred to multiple times, but it is never clearly explained in the manuscript.  For example, at line 164.  See the following sentences.

"Third, we used a mixed-effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years. While 166 meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty (Gasparrini et al., 2012; Sera et al., 2019). Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions."

It is not clear to me what exactly is meant by meta-analysis and what is being tested in the context of the time series data.  An explanation that is understandable by a general audience of potential users of this method would be helpful.  It does seem likely, as that sentence points out, that most readers would be more familiar with meta-analysis methods in the context of aggregating the results of multiple studies.  There are other helpful citations afterwards, but a more explicit connection between the familiar use case and the use case here of combining different seasonal features from a single regression model would have been helpful for reading this work.

* __Response__: Thank you for point out this ambiguity. We agree that a distinction between the more common understanding of "meta-analysis" and the one applied here needs to be made early in the manuscript.  We have added content in the introduction to clarify our methods.

     Line xyz: "Herein, meta-analysis refers to the use of seasonal features and their uncertainty estimates for trend analysis, as compared to meta-analysis for the aggregation of results from multiple studies for analysis of effect sizes or other synthesis results (i.e., Lortie et al. 2014)."

### 2021-06-24 manu_draft.Rmd
more STE revisions

Our motivating problem has several characteristics that are only partially addressed by previous methods and can further build on  GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.  However, the seasonal average within each year must be robust to inconsistent sampling times and intervals, and any trend analysis must consider the uncertainties in seasonal averages.  The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common in time series data.  This paper develops the use of GAMs with mixed-effects meta-analysis [@Gasparrini12;@Sera19] to address multi-scale trend analysis questions for which seasonal Kendall tests and the more complex STL and WRTDS methods are not well-suited.

@@

We evaluated near-surface chlorophyll (chl-a) data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20]. Sampling frequency varied somewhat over time and by station.  Every observation was included directly in the statistical models without spatial or temporal binning or averaging. Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals. Methods for back-transformation of model results are provided in the supplement.

@@

Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2). Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for January to July averages from 1991 to 2000.  Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).  The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.  Figure \@ref(fig:trndcmpex)b shows trend estimates for the same station for July to December averages from 2000 to 2010.  Unlike the first example, only the top figure shows a trend, whereas the bottom two plots do not show trends.  In both cases, only the meta-analysis results give reliable conclusions because of full propagation of uncertainty across methods. Even in cases where the p-value threshold is not of interest, the confidence intervals from the alternative methods will be inaccurate.

@@

Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01]. Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp10) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03]. Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.

@@

```{r, results = 'hide', message = F}
show_cmpplo <- function(fl, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
                        ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")")), nsim){

  load(file = fl)
  nm <- basename(fl)
  nm <- gsub('\\.RData', '', nm)
  mod <- get(nm) %>%
    pull(model) %>%
    deframe()

@@

p1 <- show_cmpplo('data/modslog_chl36.RData', doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
                  ttl =  '(a) Station 36, Jan-Jul chl. averages', ymax = NULL, nsim = nsims)
p2 <- show_cmpplo('data/modslog_chl36.RData', doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
                  ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL, nsim = nsims)
# p3 <- show_cmpplo('data/modslog_chl34.RData', doystr = 274, doyend = 364, yrstr = 2010, yrend = 2019, 
#                   ttl =  '(c) Station 34, Oct-Dec chl. averages', ymax = 15, ylab = NULL, nsim = nsims)

>>>>

Our motivating problem has several characteristics that are only partially addressed by previous methods and can further build on  GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.  However, the seasonal average within each year must be robust to inconsistent sampling times and intervals, and any trend analysis must consider the uncertainties in seasonal averages.  The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common in time series data.  This paper develops the use of GAMs with mixed-effects meta-analysis [@Gasparrini12;@Sera19] to address multi-scale trend analysis questions for which seasonal Kendall tests and the more complex STL and WRTDS methods are not well-suited. Herein, "meta-analysis" refers to the use of seasonal features and their uncertainty estimates for trend analysis, as compared to meta-analysis for the aggregation of results from multiple studies for analysis of effect sizes or other synthesis results [i.e., @Lortie14].

@@

We evaluated near-surface chlorophyll (chl-a) data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20]. Sampling frequency varied somewhat over time and by station. Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the the intended use case for the methods herein. Every observation was included directly in the statistical models without spatial or temporal binning or averaging. Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals. Methods for back-transformation of model results are provided in the supplement.

@@

Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2). Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for January to June averages from 1991 to 2000.  Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).  The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.  Figure \@ref(fig:trndcmpex)b shows trend estimates for the same station for July to December averages from 2000 to 2010.  Unlike the first example, only the top figure shows a trend, whereas the bottom two plots do not show trends.  In both cases, only the meta-analysis results give reliable conclusions because of full propagation of uncertainty across methods. Even in cases where the p-value threshold is not of interest, the confidence intervals from the alternative methods will be inaccurate.

@@

Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01]. Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp10) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.  Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test. In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].  Moreover, use of methods that fully account for uncertainty in sampling design is recommended over more conventional analyses given the certainty of conclusions in the former.  Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used. Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences.

@@

```{r, results = 'hide', message = F}
show_cmpplo <- function(datin, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
                        ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")")), nsim){

  mod <- datin %>%
    pull(model) %>%
    deframe()

@@

load(file = 'data/modslog_chl32.RData')
datin <- modslog_chl32

p1 <- show_cmpplo(datin, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
                  ttl =  '(a) Station 36, Jan-Jun chl. averages', ymax = NULL, nsim = nsims)
p2 <- show_cmpplo(datin, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
                  ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL, nsim = nsims)
# p3 <- show_cmpplo('data/modslog_chl34.RData', doystr = 274, doyend = 364, yrstr = 2010, yrend = 2019, 
#                   ttl =  '(c) Station 34, Oct-Dec chl. averages', ymax = 15, ylab = NULL, nsim = nsims)
